---
title: "Monthly report of heating usage"
author: "Hugo Pineda Hernández"
date: today
format: 
  html:
    toc: true
    self-contained: true
    smooth-scroll: true
execute:
  echo: false
  message: false
params:
  db_file: "file.sqlite"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
warning=FALSE, message = FALSE
)
```

```{r}
library(DBI)
library(tidyverse)
library(lubridate)
library(patchwork)
library(ggthemes)
theme_set(theme_minimal(14))
```

```{r}
db.file <- params$db_file
```

```{r}
thermostat_db <- DBI::dbConnect(RSQLite::SQLite(), db.file)
```

```{r}
# collect tables from db
datehour <- tbl(thermostat_db, "datehour") |> 
    collect() 

thermostat <- tbl(thermostat_db, "thermostat") |> 
    collect() |> 
  mutate(
    datetime = ymd_hms(datetime)
  )

weather <- tbl(thermostat_db, "weather") |> 
    collect() 

DBI::dbDisconnect(thermostat_db)
```

```{r}
# join thermostat with date data
thermostat.d <- thermostat |> 
  inner_join(datehour, by = "datehour_id") 
# join with weather time
thermostat.weather <- thermostat.d |> 
  inner_join(weather, by = "datehour_id")
```

## Active heating time

```{r}
# get the number of hours the heating was on per month an year
thermostat.per.month <- thermostat.d |> 
  group_by(year, month) |> 
  summarise(
    time_on_h = sum(boiler_on) / 60 / 60
  ) |> 
  ungroup()
```

The next figure shows the number of hours of active heating time per month and year.

```{r}
thermostat.per.month |> 
  mutate(
    month = month(month, label = TRUE)
  ) |> 
  ggplot(aes(x = month, y = time_on_h)) +
  geom_col(width = 0.5) +
  labs(
    x = "Month",
    y = "Hours"
  ) +
  theme(
    panel.grid.major.x = element_blank()
  ) +
  facet_wrap(~year)

```

The next figure compares the number of hours of active heating time per month between different years.

```{r}
thermostat.per.month |> 
  mutate(
    month = month(month, label = TRUE)
  ) |> 
  ggplot(aes(x = month, y = time_on_h, fill = as.factor(year))) +
  geom_col(position = "dodge") +
  labs(
    x = "Month",
    y = "Hours",
    fill = "Year"
  ) +
  theme(
    panel.grid.major.x = element_blank()
  ) +
  scale_fill_colorblind()
```

## Weather data

If the thermostat setting are constant, the heating time will depend the outside temperature. 

```{r}
thermostat.temp.per.day <- thermostat.weather |> 
  group_by(year, month, day) |> 
  summarise(
    time_on_h = sum(boiler_on) / 60 / 60,
    ext_temperature = mean(temperature, na.rm = T)
  ) |> 
  ungroup() |> 
  mutate(
    dateday = ymd(str_c(year, month, day))
  )
```

The next figures compare the number of active heating hours with the average temperature outside:

```{r}
thermostat.temp.per.day |>
  ggplot(aes(x = ext_temperature, y = time_on_h)) +
  geom_point(size = 4, alpha = 0.5) +
   labs(
    x ="Average outside temperature",
    y = "Hours per day",
    title = "Active heating vs outside temperature",
    subtitle = "Per year and month"
    
  ) +
  scale_x_continuous(
    labels = scales::label_number(suffix = "\u00b0C")
    ) +
  scale_color_colorblind() +
  facet_grid(year~month)
```

```{r}
thermostat.temp.per.day |>
  ggplot(aes(x = ext_temperature, y = time_on_h, color = as.factor(year))) +
  geom_point(size = 4, alpha = 0.5) +
   labs(
    x ="Average outside temperature",
    y = "Hours per day",
    color = "Year",
    title = "Active heating vs outside temperature",
    subtitle = "Data points colored by year"
    
  ) +
  scale_x_continuous(
    labels = scales::label_number(suffix = "\u00b0C")
    ) +
  scale_color_colorblind()
```


